Symmetric Circuits for Rank Gates

### 2018-02-16 paper.tex
Updated first four sections.
I have realised that my generalisation of symmetry is flawed and will need an overhaul. However, I want to keep a copy so I'm updating git.
This version is the told one where we are just speaking of $\mathcal{D}$-invarient functions.
13-Nov-2017
editing process
24th January 2018
February 5 2018
New introduction. Also changes to top-level file to put in Section labels. And a new reference.

% This first part of the file is called the PREAMBLE. It includes customizations
% and command definitions. The preamble is everything between \documentclass
% and \begin{document}.

\usepackage[margin=1in]{geometry} % set the margins to 1in on all sides
\usepackage{graphicx} % to include figures
\usepackage{amsmath} % great math stuff
\usepackage{amsfonts} % for blackboard bold, etc
\usepackage{amsthm} % better theorem environments
\usepackage{xspace}
\usepackage{subfiles}%allow for subfiles
\usepackage{blindtext}
\usepackage{mymacros}

@@

\section{Background}
\subfile{sections/background}
\newpage

\section{Symmetric Circuits}
\subfile{sections/symmetric-circuits}
\newpage

\section{Symmetry and Support}
\subfile{sections/symmetry-and-support}
\newpage

\section{Translating Circuits into Formulas}
\subfile{sections/translating-circuits-into-formulas}
\newpage

\section{Concluding Remarks}
\subfile{sections/conclusion}
\newpage

\bibliographystyle{plain} \bibliography{references}
\end{document}

>>>>

% This first part of the file is called the PREAMBLE. It includes customizations
% and command definitions. The preamble is everything between \documentclass
% and \begin{document}.asd

\usepackage[margin=1in]{geometry} % set the margins to 1in on all sides
\usepackage{graphicx} % to include figures
\usepackage{amsmath} % great math stuff
\usepackage{amsfonts} % for blackboard bold, etc
\usepackage{amsthm} % better theorem environments
\usepackage{xspace}
\usepackage{subfiles}%allow for subfiles
\usepackage{blindtext}
\usepackage{mymacros}
\usepackage{anujmacros}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{etoolbox}

\newenvironment{itemizens}
{ \begin{itemize}
    \setlength{\itemsep}{0pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}     }
{ \end{itemize}                  }

@@

\section{Background}\label{sec:background}
\subfile{sections/background}
\newpage

\section{Symmetric Circuits}\label{sec:symm-circ}
\subfile{sections/symmetric-circuits}
\newpage

\section{Symmetry and Support}\label{sec:symm-support}
\subfile{sections/symmetry-and-support}
\newpage

\section{Transparent Circuits}\label{sec:transparent}
\subfile{sections/transparency-and-unique-extensions.tex}

\section{Translating Formulas into Circuits}\label{sec:formulas-to-circuits}
\subfile{sections/translating-formulas-to-cicuits.tex}
\newpage

\section{Translating Circuits into Formulas}\label{sec:circuits-to-formulas}
\subfile{sections/translating-circuits-into-formulas}
\newpage

\section{Concluding Remarks}
\subfile{sections/conclusion}
\newpage
\bibliographystyle{plain} \bibliography{references}
\end{document}

### 2018-02-16 sections/introduction.tex
New introduction. Also changes to top-level file to put in Section labels. And a new reference.

\documentclass[../paper.tex]{subfiles}
\begin{document}
When we study algorithmic complexity we often study complexity classes defined
in terms of some restricted class of machines. Circuit complexity, where one
makes use of similarly restricted families of Boolean circuits rather than
machines, has long served as an alternative approach. These circuits models have
a natural attraction for researchers, as their simpler structure seems to make
them more amendable to analysis and their connections with machine models
provides an alternative avenue for better understanding fundamental problems in
complexity theory, including the $\PT$ vs $\NP$ problem.

>>>>

\documentclass[../paper.tex]{subfiles}
\begin{document}

The study of extensions of fixed-point logics plays an important role
in the field of descriptive complexity theory.  In particular,
fixed-point logic with counting ($\FPC$) has become a reference logic
in the search for a logic for polynomial-time
(see~\cite{Dawar-siglog}).  In this context, Anderson and
Dawar~\cite{AndersonD17} provide an interesting characterization of
the expressive power of $\FPC$ in terms of circuit complexity.  They
show that the properties expressible in this logic are exactly those
that can be decided by polynomially-uniform families of circuits (with
threshold gates) satisfying a natural \emph{symmetry} condition.  Not
only does this illustrate the robustness of $\FPC$ as a complexity
class within $\PT$ by giving a distinct and natural characterization
of it, it also demonstrates that the techniques for proving
inexpressibility in the field of finite model theory can be understood
as lower-bound methods against a natural circuit complexity class.
This raises an obvious question (explicitly posed in the concluding
section of~\cite{AndersonD17}) of how to obtain circuit
characterizations of logics more expressive than $\FPC$, such as
choiceless polynomial time (CPT) and fixed-point logic with rank
($\FPR$).  It is this last question that we address in this paper.

Fixed-point logic with rank extends the expressive power of $\FPC$ by
means of operators that allow us to define the rank of a matrix over a
finite field.  Such operators are natural extensions of
counting---counting the dimension of a definable vector space
rather than just the size of a definable set.  At the same time they
make the logic rich enough to express many of the known examples that
separate $\FPC$ from $\PT$.  Rank logics were first introduced
in~\cite{Dawar09logicswith}.  The version $\FPR$ we consider here is that
defined by Gr\"adel and Pakusa~\cite{GradelP15a} where the prime
characteristic is a parameter to the rank operator, and we do not have
a distinct operator for each prime number.  Formal definitions of
these logics are given in Section~\ref{sec:background}.  We give a circuit
characterization, in terms of symmetric circuits, of $\FPR$.  One
might think, at first sight, that this is a simple matter of extending
the circuit model with gates for computing the rank of a matrix.
It turns out, however, that the matter is not so simple as the
symmetry requirement interacts in surprising ways with such rank
gates.  It requires a new framework for defining classes of such
circuits, which yields remarkable new insights.

It is important to clarify at the outset what is meant by symmetric
circuits.  Indeed, there are different senses in which the word
\emph{symmetry} is used in the context of circuits (and also in this
paper).  We say that a Boolean function $f:\{0,1\}^n \ra \{0,1\}$ is
symmetric if the value of the function on a string $s$ is determined
by the number of $1$s in $s$.  In other words, $f$ is invariant under
\emph{all} permutations of its input.  In contrast, when we consider
the input to a Boolean function to be the adjacency matrix of an
$n$-vertex graph, for example, and
$f : \{0,1\}^{n \choose 2} \ra \{0,1\}$ decides a graph property, then
$f$ is invariant under all permutations of its input induced by
permutations of the $n$ vertices of the graph.  We call such a
function \emph{graph-invariant}.  More generally, for a relational
vocabulary $\tau$ and a standard encoding of $n$-element
$\tau$-structures as strings over $\{0,1\}$, we can say that function
taking such strings as input is $\tau$-invariant if it is invariant
under permutations induced by the $n$ elements.
A circuit $C$ computing such an invariant function is said to be \emph{symmetric} if every
permutation of the $n$ elements extends to an automorphism of $C$.  It
is families of symmetric circuits in this sense that characterize
$\FPC$ in~\cite{AndersonD17}.  The restriction to symmetric circuits
arises naturally in the study of logics and has appeared previously
under the names of generic circuits in the work
of~\cite{DENENBERG1986216} and explicitly order-invariant circuits in
the work of Otto~\cite{Otto1997}.

The main result of~\cite{AndersonD17} says that the properties of
$\tau$-structures definable in $\FPC$ are exactly those that can be
decided by $\PT$-uniform families of symmetric circuits using AND, OR,
NOT and majority gates.  Note that each of these gates itself computes
a Boolean function that is symmetric in the strong sense identified
above.  On the other hand, a gate for computing a rank threshold
function, e.g.\ one that takes as input a $n \times n$ matrix and
outputs $1$ if the rank of the matrix is greater than a threshold $t$,
is \emph{not} symmetric.  In our circuit characterization of $\FPR$ we
necessarily have to consider such non-symmetric gates.  Indeed, we
show in Section~\ref{sec:symm-circ} that  $\PT$-uniform families of
symmetric circuits using gates for \emph{any} symmetric functions do
not take us beyond the power of $\FPC$.  This is a further
illustration of the robustness of $\FPC$.  In order to go beyond it,
we need to introduce gates for Boolean functions that are not
symmetric.  We construct a systematic framework for including
functions computing $\tau$-invariant functions for arbitrary
multi-sorted relational vocabularies $\tau$ in
Section~\ref{sec:symm-circ}.  We also explore what it means for such
circuits to be symmetric.

The proof of the circuit characterization of $\FPC$ relies on the
\emph{support theorem} proved in~\cite{AndersonD17}.  This establishes
that for any $\PT$-uniform family of circuits using AND, OR, NOT and
majority gates there is a constant $k$ such that every gate has a
support of size at most $k$.  That is to say that we can associate
with every gate $g$ in the circuit $C_n$ (the circuit in the family
that works on $n$-element structures) a subset $X$ of $[n]$ of size at
most $k$ such that any permutation of $[n]$ fixing $X$ pointwise
extends to an automorphism of $C_n$ that fixes $g$.  This theorem is
crucial to the translation of the family of circuits into a formula of
$\FPC$, which is the difficult (and novel) direction of the
equivalence.  In attempting to do the same with circuits that now use
rank-threshold gates we are faced with the difficulty that the proof
of the support theorem in~\cite{AndersonD17} relies in an essential
way on the fact that the Boolean function computed at each gate is
symmetric.  We are able to overcome this difficulty and prove a
support theorem for circuits with rank gates but this requires
substantial, novel technical machinery, which we develop in
Section~\ref{sec:symm-support}.

Another crucial ingredient in the proof of Anderson and Dawar is that
we can eliminate redundancy in the circuit $C_n$ by making it
\emph{rigid}.  That is, we can ensure that the \emph{only}
automorphisms of $C_n$ are those that are induced by permutations of
$[n]$.  Here we face the difficulty that identifying the symmetries
and eliminating redundancy in a circuit that involves gates computing
$\tau$-invariant functions requires us to solve the isomorphism
problem for $\tau$-structures.  This is a hard problem (or, at least,
one that we do not know how to solve efficiently) even when the
$\tau$-structures are $0$-$1$-matrices.  We overcome this difficulty
by placing a further restriction on circuits that we call
\emph{transparency}.  Circuits satisfying this condition have the
property that their lack of redundancy is transparent.  We explore
this condition and its consequences in Section~\ref{sec:transparent}.

In the characterization of $\FPC$, the translation from formulas into
families of circuits is easy and, indeed, standard.  In our case, we
have to show that formulas of $\FPR$ translate into uniform families of
circuits using rank-threshold gates that are symmetric and
transparent.  This is somewhat more involved technically and the
details are presented in Section~\ref{sec:formulas-to-circuits}.
Finally, with all the tools in place, the translation of such
$\PT$-uniform families of circuits into formulas of $\FPR$ given in Section~\ref{sec:circuits-to-formulas} completes
the characterization.

\end{document}

When we study algorithmic complexity we often study complexity classes defined
in terms of some restricted class of machines. Circuit complexity, where one
makes use of similarly restricted families of Boolean circuits rather than
machines, has long served as an alternative approach. These circuits models have
a natural attraction for researchers, as their simpler structure seems to make
them more amendable to analysis and their connections with machine models
provides an alternative avenue for better understanding fundamental problems in
complexity theory, including the $\PT$ vs $\NP$ problem.
